---
title: "Team 1 Mid-term Project"
author: "Team 1 (Names: Adam Kritz; Huang He..."
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---



```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes a lot of the helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
library(ezids)
```

# Midterm Project  

## Importing the Data  

```{r setup, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

```{r read the dataset}
diamonds = data.frame(read.csv("diamonds.csv"))
summary(diamonds)
str(diamonds)
diamonds$color = factor(diamonds$color, order=T, levels = c('D', 'E', 'F', 'G', 'H', 'I', 'J'))
diamonds$clarity = factor(diamonds$clarity, order=T, levels = c('I1', 'SI2', 'SI1', 'VS2', 'VS1', 'VVS2', 'VVS1', 'IF'))
diamonds$cut = factor(diamonds$cut, order=T, levels = c('Fair', 'Good', 'Very Good', 'Premium', 'Ideal'))
colnames(diamonds)[1] = "ID"
colnames(diamonds)
summary(diamonds)
```

## Diamonds of all price  

### Distribution of each variables  

#### price  
```{r}
library(ggplot2)
library(ezids)
# histogram
p = ggplot(diamonds,aes(x=price)) + 
  geom_histogram(breaks = seq(300,18900,100),fill = "red",color = "grey",alpha = 0.3) + labs(title = "Distribution of Price", x = "Price",y = "frequency")
p  + theme_bw()
# boxplot
p = ggplot(diamonds,aes(y=price)) + 
  geom_boxplot(colour="orange", fill="#7777cc", outlier.colour="red", outlier.shape=8, outlier.size=4) + 
  labs(title = "Distribution of Price")
p + theme_bw() + theme(panel.grid=element_blank())
# QQ-plot
qqnorm(diamonds$price, main="Q-Q plot of Price", col = 'blue', ylab = 'Price') 
qqline(diamonds$price)

```


#### carat  
```{r}
# histogram
p = ggplot(diamonds,aes(x=carat)) + 
  geom_histogram(breaks = seq(0.2,5.1,0.1),fill = "red",color = "black",alpha = 0.3) + labs(title = "Distribution of Carat", x = "Carat",y = "frequency")
p  + theme_bw()
# boxplot
p = ggplot(diamonds,aes(y=carat)) + 
  geom_boxplot(colour="orange", fill="#7777cc", outlier.colour="red", outlier.shape=8, outlier.size=4) + 
  labs(title = "Distribution of Carat")
p + theme_bw() + theme(panel.grid=element_blank())
# QQ-plot
qqnorm(diamonds$carat, main="Q-Q plot of Carat", col = 'blue', ylab = 'Carat') 
qqline(diamonds$carat)
```

#### table  
```{r}
# histogram
p = ggplot(diamonds,aes(x=table)) + 
  geom_histogram(breaks = seq(42,96,1),fill = "red",color = "black",alpha = 0.3) + labs(title = "Distribution of Table", x = "Table",y = "frequency")
p  + theme_bw()
# boxplot
p = ggplot(diamonds,aes(y=table)) + 
  geom_boxplot(colour="orange", fill="#7777cc", outlier.colour="red", outlier.shape=8, outlier.size=4) + 
  labs(title = "Distribution of Table")
p + theme_bw() + theme(panel.grid=element_blank())
# QQ-plot
qqnorm(diamonds$table, main="Q-Q plot of Table", col = 'blue', ylab = 'Table') 
qqline(diamonds$table)
```

#### depth  
```{r}
# histogram
p = ggplot(diamonds,aes(x=depth)) + 
  geom_histogram(breaks = seq(43,79,0.5),fill = "red",color = "black",alpha = 0.3) + labs(title = "Distribution of Depth", x = "depth",y = "frequency")
p  + theme_bw()
# boxplot
p = ggplot(diamonds,aes(y=depth)) + 
  geom_boxplot(colour="orange", fill="#7777cc", outlier.colour="red", outlier.shape=8, outlier.size=4) + 
  labs(title = "Distribution of Depth")
p + theme_bw() + theme(panel.grid=element_blank())
# QQ-plot
qqnorm(diamonds$depth, main="Q-Q plot of Depth", col = 'blue', ylab = 'Depth') 
qqline(diamonds$depth)
```


#### cut  
```{r}
barplot(table(diamonds$cut), col = c(2, 3, 4, 7, 6), main = 'Barplot of Cut', xlab = "Cut", ylab = "Frequency")
```

#### color  
```{r}
barplot(table(diamonds$color), col = c(2, 3, 4, 7, 6, 8, 9), main = 'Barplot of Color', xlab = "Color", ylab = "Frequency")
```


#### clarity  
```{r}
barplot(table(diamonds$clarity), col = c(2, 3, 4, 7, 6, 8, 9, 5), main = 'Barplot of Color', xlab = "Clarity", ylab = "Frequency")
```

### Relationship between variables  

#### SMART:what variable effects price ?  
```{r results='markup'}
# matrix and corrplot
diamonds_num = diamonds[,c(2,6,7,8,9,10,11)]
mtx_num = cor(diamonds_num)
xkabledply(mtx_num,title = "Matrix of All Numerical Variables")
library("corrplot")
corrplot(mtx_num,method = "square", type="upper")
```

#SMART: Does cut methods effect price?  
```{r}
p = ggplot(diamonds,aes(x = cut,y=price)) + 
  geom_boxplot(colour="grey", fill=c(2, 3, 4, 7, 6), outlier.colour="red", outlier.shape=8, outlier.size=4) + 
  labs(title = "Distribution of Price in Different Cut methods")
p + theme_bw() + theme(panel.grid=element_blank())
anv_pri_cut = aov(price ~ cut,data = diamonds)
xkabledply(anv_pri_cut,title = "Cut vs Price ANOVA result summary")

```

```{r results='markup'}
tukey_pri_cut <- TukeyHSD(anv_pri_cut)
tukey_pri_cut
```
Diamonds with different cut methods do have different price averages.  

#SMART: Does clarity levels effect price?  
```{r}
p = ggplot(diamonds,aes(x = clarity,y=price)) + 
  geom_boxplot(colour="orange", fill=c(2, 3, 4, 7, 6 ,5 ,8 ,9), outlier.colour="red", outlier.shape=8, outlier.size=4) + 
  labs(title = "Distribution of Price in Different Clarity levels")
p + theme_bw() + theme(panel.grid=element_blank())
anv_pri_cla = aov(price ~ clarity,data = diamonds)
xkabledply(anv_pri_cla,title = "Clarity vs Price ANOVA result summary")
```

```{r results='markup'}
tukey_pri_cla <- TukeyHSD(anv_pri_cla)
tukey_pri_cla
```
Diamonds in different clarity levels do have different price averages.  

#SMART: Does color effect price?  
```{r}
p = ggplot(diamonds,aes(x = color,y=price)) + 
  geom_boxplot(colour="orange", fill=c(2, 3, 4, 7, 6 ,5 ,8 ), outlier.colour="red", outlier.shape=8, outlier.size=4) + 
  labs(title = "Distribution of Price in Different Color")
p + theme_bw() + theme(panel.grid=element_blank())
anv_pri_col = aov(price ~ color,data = diamonds)
xkabledply(anv_pri_col,title = "Clarity vs Price ANOVA result summary")
```


```{r results='markup'}
tukey_pri_col <- TukeyHSD(anv_pri_col)
tukey_pri_col
```

Diamonds in different clarity levels do have different price averages.  


#### what variable effects cutting methods?  

#SMART: Do different ranks of color have different cutting methods?  

```{r results='markup'}
tab_cut_col = table(diamonds$color,diamonds$cut)
chitest = chisq.test(tab_cut_col)
chitest
```

```{r}
names = colnames(tab_cut_col)
cols = c("#ED1C24","#22B14C","#FFC90E","#3f48CC","pink")
piepercent_d = paste(names,":",round(100*tab_cut_col[1,]/sum(tab_cut_col[1,])), "%")
piepercent_e = paste(names,":",round(100*tab_cut_col[2,]/sum(tab_cut_col[2,])), "%")
piepercent_f = paste(names,":",round(100*tab_cut_col[3,]/sum(tab_cut_col[3,])), "%")
piepercent_g = paste(names,":",round(100*tab_cut_col[4,]/sum(tab_cut_col[4,])), "%")
piepercent_h = paste(names,":",round(100*tab_cut_col[5,]/sum(tab_cut_col[5,])), "%")
piepercent_i = paste(names,":",round(100*tab_cut_col[6,]/sum(tab_cut_col[6,])), "%")
piepercent_j = paste(names,":",round(100*tab_cut_col[7,]/sum(tab_cut_col[7,])), "%")
par(mfrow=c(2,2))
pie(tab_cut_col[1,], labels=piepercent_d, main = "D rank of color", col=cols)
pie(tab_cut_col[2,], labels=piepercent_e, main = "E rank of color", col=cols)
pie(tab_cut_col[3,], labels=piepercent_f, main = "F rank of color", col=cols)
pie(tab_cut_col[4,], labels=piepercent_g, main = "G rank of color", col=cols)
pie(tab_cut_col[5,], labels=piepercent_h, main = "H rank of color", col=cols)
pie(tab_cut_col[6,], labels=piepercent_i, main = "I rank of color", col=cols)
pie(tab_cut_col[7,], labels=piepercent_j, main = "J rank of color", col=cols)
```
Different ranks of color have different cutting methods.  



#SMART: Do different ranks of clarity have different cutting methods?  
```{r results='markup'}
tab_cut_cla = table(diamonds$clarity,diamonds$cut)
chitest = chisq.test(tab_cut_cla)
chitest
```

```{r}
names = colnames(tab_cut_cla)
cols = c("#ED1C24","#22B14C","#FFC90E","#3f48CC","pink")
piepercent_1 = paste(names,":",round(100*tab_cut_cla[1,]/sum(tab_cut_cla[1,])), "%")
piepercent_2 = paste(names,":",round(100*tab_cut_cla[2,]/sum(tab_cut_cla[2,])), "%")
piepercent_3 = paste(names,":",round(100*tab_cut_cla[3,]/sum(tab_cut_cla[3,])), "%")
piepercent_4 = paste(names,":",round(100*tab_cut_cla[4,]/sum(tab_cut_cla[4,])), "%")
piepercent_5 = paste(names,":",round(100*tab_cut_cla[5,]/sum(tab_cut_cla[5,])), "%")
piepercent_6 = paste(names,":",round(100*tab_cut_cla[6,]/sum(tab_cut_cla[6,])), "%")
piepercent_7 = paste(names,":",round(100*tab_cut_cla[7,]/sum(tab_cut_cla[7,])), "%")
piepercent_8 = paste(names,":",round(100*tab_cut_cla[8,]/sum(tab_cut_cla[8,])), "%")
par(mfrow=c(2,2))
pie(tab_cut_cla[1,], labels=piepercent_1, main = "I1 rank of color", col=cols)
pie(tab_cut_cla[2,], labels=piepercent_2, main = "SI2 rank of color", col=cols)
pie(tab_cut_cla[3,], labels=piepercent_3, main = "SI1 rank of color", col=cols)
pie(tab_cut_cla[4,], labels=piepercent_4, main = "VS2 rank of color", col=cols)
pie(tab_cut_cla[5,], labels=piepercent_5, main = "VS1 rank of color", col=cols)
pie(tab_cut_cla[6,], labels=piepercent_6, main = "VVS2 rank of color", col=cols)
pie(tab_cut_cla[7,], labels=piepercent_7, main = "VVS1 rank of color", col=cols)
pie(tab_cut_cla[8,], labels=piepercent_8, main = "IF rank of color", col=cols)
```
Different ranks of clarity have different cutting methods.  

```{r}

```


```{r}

```


```{r}

```


```{r}

```




